Tracing the Meta-Level: PyPy's Tracing JIT Compiler

### 2009-04-03 talk/icooolps2009/paper.tex
it's not simple any more
Some more references. Finally take a stab at the benchmark section.

A recent approach to getting better performance for dynamic languages is that of
tracing JIT compilers \cite{XXX}. Writing a tracing JIT compiler is relatively
simple. It can be added to an existing interpreter for a language,
the interpreter takes over some of the functionality of the compiler and
the machine code generation part can be simplified.

@@

Tracing JITs are an idea initially explored by the Dynamo project
\cite{bala_dynamo:transparent_2000} in the context of dynamic optimization of
machine code at runtime. The techniques were then successfully applied to Java
VMs \cite{gal_hotpathvm:effective_2006}. It also turned out that they are a
relatively simple way to implement a JIT compiler for a dynamic language
\cite{mason_chang_efficient_2007}. The technique is now
being used by both Mozilla's TraceMonkey JavaScript VM \cite{XXX} and Adobe's
Tamarin ActionScript VM \cite{XXX}.

@@

\textbf{Assembler Backends:} The tracing interpreter uses a well-defined
interface to an assembler backend for code generation. This makes it possible to
easily port the tracing JIT to various architectures (including, we hope, to
virtual machines such as the JVM where backend could generate bytecode at
runtime). At the moment the only implemented backend is a simple 32-bit
Intel-x86 backend.

@@

In this section we try to evaluate the work done so far by looking at some
benchmark numbers. Since the work is not finished, these benchmarks can only be
preliminary. All benchmarking was done on a machine with a 1.4 GHz Pentium M
processor and 1GiB RAM, using Linux 2.6.27.

>>>>

A recent approach to getting better performance for dynamic languages is that of
tracing JIT compilers \cite{gal_hotpathvm:effective_2006,
mason_chang_efficient_2007}. Writing a tracing JIT compiler is relatively
simple. It can be added to an existing interpreter for a language,
the interpreter takes over some of the functionality of the compiler and
the machine code generation part can be simplified.

@@

Tracing JITs are an idea initially explored by the Dynamo project
\cite{bala_dynamo:transparent_2000} in the context of dynamic optimization of
machine code at runtime. The techniques were then successfully applied to Java
VMs \cite{gal_hotpathvm:effective_2006}. It also turned out that they are a
relatively simple way to implement a JIT compiler for a dynamic language
\cite{mason_chang_efficient_2007}. The technique is now
being used by both Mozilla's TraceMonkey JavaScript VM
\cite{andreas_gal_trace-based_2009} and has been tried for Adobe's Tamarin
ActionScript VM \cite{chang_tracing_2009}.

@@

\textbf{Assembler Backends:} The tracing interpreter uses a well-defined
interface to an assembler backend for code generation. This makes it possible to
easily port the tracing JIT to various architectures (including, we hope, to
virtual machines such as the JVM where backend could generate bytecode at
runtime). At the moment the only implemented backend is a 32-bit
Intel-x86 backend.

@@

In this section we try to evaluate the work done so far by looking at some
benchmark numbers. Since the work is not finished, these benchmarks can only be
preliminary. All benchmarking was done on an otherwise idle machine with a 1.4
GHz Pentium M processor and 1GiB RAM, using Linux 2.6.27.

The first round of benchmarks (Figure \ref{fig:bench1}) are timings of the
example interpreter (Figure \ref{fig:tlr-basic}) used in this paper computing
the square of 46340 (the smallest number so that the square still fits into a 32
bit word) using the bytecode of Figure \ref{fig:square}. The results for various
constellations are as follows:

\begin{enumerate}
\item The interpreter translated to C without any JIT inserted at all.
\item The tracing JIT is enabled, but no interpreter-specific
hints are applied. This corresponds to the trace in Figure
\ref{fig:trace-normal}. The time includes the time it takes to trace and the
production of the machine code, as well as the fallback interpreter to leave the
machine code. The threshold when to consider a loop to be hot is 40 iterations.
\item The hints as in Figure \ref{fig:tlr-full} are applied, which means the loop of
the square function is reflected in the trace. Constant folding of green
variables is disabled though. This corresponds to the trace in Figure
\ref{fig:trace-no-green-folding}. XXX
\item Same as before, but with constant folding enabled. This corresponds to the
trace in Figure \ref{fig:trace-full}. This speeds up the square function nicely,
making it about six times faster than the pure interpreter.
\item Same as before, but with the threshold set so high that the tracer is
never invoked. This measures the overhead of the profiling. For this interpreter
the overhead seems rather large, with 50\% slowdown du to profiling. This is
because the example interpreter needs to do one hash table lookup per loop
iteration. For larger interpreters (e.g. the Python one) it seems likely that
the overhead is less significant, given that many operations in Python need
hash-table lookups themselves.
\item Runs the whole computation on the tracing interpreter for estimating the
involved overheads of tracing. The trace is not actually recorded (which would be a
memory problem), so in reality the number is even higher. Due to the double
interpretation, the overhead is huge. It remains to be seen whether that will be
a problem for practical interpreters.
\item For comparison, the time of running the interpreter on top of CPython
(version 2.5.2).
\end{enumerate}

\begin{figure}
\noindent
\begin{tabular}{|l|r|}
\hline
 &ratio\tabularnewline
\hline
Interpreter compiled to C, no JIT &1\tabularnewline \hline
Normal Trace Compilation &1.20\tabularnewline \hline
Unfolding of Language Interpreter Loop &XXX\tabularnewline \hline
Full Optimizations &0.17\tabularnewline \hline
Profile Overhead &1.51\tabularnewline \hline
Interpreter run by Tracing Interpreter &860.20\tabularnewline \hline
Interpreter run by CPython &256.17\tabularnewline \hline
\end{tabular}
\label{fig:bench1}
\caption{Benchmark results of example interpreter computing the square of
46340}
\end{figure}
